---
title: "Missing Data"
#author: "Marta Fernandes"
output:
  html_document:
    fig.align: center
    fig_caption: yes
    fig_height: 5
    fig_width: 9
    highlight: tango
    theme: united
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Source MIMIC-III demo version

Go to this site: https://mimic.physionet.org/gettingstarted/demo/

In the page, follow the steps in "Accessing the demo data".


## Extract data and set working directory

1. SQL queries:

Extract the datasets with each one of the following lines of code in SQL:

```{r message=FALSE, warning=FALSE}
## SELECT * FROM admissions 
## SELECT * FROM icustays
## SELECT * FROM chartevents
```

Save admissions.csv, icustays.csv and chartevents.csv in your desktop folder: "MIMIC_data_files" 

Observation: In case you will not use SQL, you can simply download the tables from the source MIMIC-III demo version above mentioned.

2. Set working directory and read data files into R:

```{r message=FALSE, warning=FALSE}

# Set your working directory
work_dir <- "C:/Users/Marta/Desktop/MIMIC_data_files" # here your directory

setwd(work_dir)

# These are the tables for this workshop:
adm <- read.csv(file="admissions.csv", header=TRUE, sep =",")
icu <- read.csv(file="icustays.csv", header=TRUE, sep =",")
icu_adm  <- merge(adm, icu, by = c("SUBJECT_ID", "HADM_ID"))
vitals  <-read.csv(file="chartevents.csv", header=TRUE, sep =",")


```

## Install the missing data packages

Uncomment the installation of the packages, if running the file for the first time.

```{r message=FALSE, warning=FALSE}

# install mice
#install.packages("mice")
library(mice)

#install.packages("VIM")
library(VIM)

#install.packages("Amelia")
library(Amelia)

```

## Exercise: Selection of variables

Aim: Select variables to analyze missing values.


### Vital Signs

D_ITEMS is sourced from two distinct ICU databases. The main consequence is that there are duplicate ITEMID for each concept. For example, heart rate is captured both as an ITEMID of 211 (CareVue) and as an ITEMID of 220045 (Metavision). As a result, it is necessary to search for multiple ITEMID to capture a single concept across the entire database.

For more information read: https://mimic.physionet.org/mimictables/d_items/

In GitHub you may find code regarding vital signs and respective items IDs:
  
https://github.com/MIT-LCP/mimic-code/blob/master/concepts/firstday/vitals-first-day.sql


We will use data from chartevents, namely heart rate (hr) and pulse oximetry (spO2) measurements. Attention to the units. Since much of the information is manually typed in the system, human error can always be present.

Let's see the amount of missing values in both vitals.


```{r message=FALSE, warning=FALSE}

hr <- vitals[((vitals$ITEMID = 211) | (vitals$ITEMID = 220045)) & (vitals$VALUEUOM == "bpm"),]

summary(hr$VALUENUM) 

sp <- vitals[((vitals$ITEMID = 646) | (vitals$ITEMID = 220277)) & (vitals$VALUEUOM == "%"),]

summary(sp$VALUENUM) 


```

We observe there are no missing values for heart rate and 18 "NA" for spO2. Heart rate has a higher frequency of measurement, (however this is a demo version of the database).

##  Exercise: Handling missing data in dataframes 

Two of the simplest methods to handle missing data are following presented. Recoding a missing value consists in assigning a value to an already existing value (eg. outlier) which we want to recode as missing. Excluding missing values can be performed by excluding objects (patients, rows in the dataframe) or variables (columns in the dataframe) with significant amount of missing data (see this chapter contents for more detailed theory).

### Outliers

Before handling missing data, we should analyze the presence of outliers (refer to the respective chapter in the course). If we apply an imputation method before processing outliers, our imputation will be based on incorrect data and therefore not valid. 

Looking at the distribution of heart rate for all patients, we can assess potential outliers through a box plot.

Let's observe the distribution of heart rate for all the patients in the dataset.

```{r message=FALSE, warning=FALSE}

#install.packages("plotly")

library(plotly)

p <- plot_ly(y = ~hr$VALUENUM,  name = "Heart rate", type = "box") %>%
layout(
    title = "Heart rate distribution for all patients",
    yaxis = list(title = 'Value [bpm]')
    ) 
p

```

We observe there are a few outliers. The value 300 bpm is still a possible value to exist, however 580 bpm is clearly an outlier and should be removed.

### Missing data recoding

Aim: Recode a value as missing.

We can recode all values outside the physiological ranges [0, 300] bpm as "NA" in the dataset for all patients.

```{r message=FALSE, warning=FALSE}

hr$VALUENUM[hr$VALUENUM > 300] <- NA

hr$VALUENUM[hr$VALUENUM < 0] <- NA

summary(hr$VALUENUM) 

```

We observe we have now a missing value "NA", correspondent to value 580 bpm, which is no longer in the dataset.

The distribution of heart rate is now within the physiological ranges.

```{r message=FALSE, warning=FALSE}

#install.packages("plotly")

library(plotly)

p <- plot_ly(y = ~hr$VALUENUM,  name = "Heart rate", type = "box") %>%
layout(
    title = "Heart rate distribution for all patients with outlier processing",
    yaxis = list(title = 'Value [bpm]')
    ) 
p

```


Let's repeat the process for pulse oximetry (SpO2). 

Let's observe the distribution of spO2 for all the patients in the dataset.

```{r message=FALSE, warning=FALSE}

#install.packages("plotly")

library(plotly)

p <- plot_ly(y = ~sp$VALUENUM,  name = "Pulse oxymetry", type = "box") %>%
layout(
    title = "Pulse oximetry distribution for all patients",
    yaxis = list(title = 'Value [%]')
    ) 
p

```

We observe there are a few outliers. Values above 100% are not possible for this variable.

We can recode all values outside the physiological ranges [0, 100] % as "NA" in the dataset for all patients.

```{r message=FALSE, warning=FALSE}

sp$VALUENUM[sp$VALUENUM > 100] <- NA

summary(sp$VALUENUM) 

```

We observe we have now 21 missing values "NA".

The distribution of pulse oximetry is now within the physiological ranges.

```{r message=FALSE, warning=FALSE}

#install.packages("plotly")

library(plotly)

p <- plot_ly(y = ~sp$VALUENUM,  name = "Pulse oxymetry", type = "box") %>%
layout(
    title = "Pulse oximetry distribution for all patients with outlier > 100 % processing",
    yaxis = list(title = 'Value [%]')
    ) 
p

```


We observe that there are several values bellow 80%. Several of these values may correspond to other variables, such as respiratory or heart rate that were mistyped in the system. For our analysis we will recoded them as "NA".

```{r message=FALSE, warning=FALSE}

sp$VALUENUM[sp$VALUENUM < 80] <- NA

summary(sp$VALUENUM) 

```


```{r message=FALSE, warning=FALSE}

#install.packages("plotly")

library(plotly)

p <- plot_ly(y = ~sp$VALUENUM,  name = "Pulse oxymetry", type = "box") %>%
layout(
    title = "Pulse oximetry distribution for all patients with outlier processing",
    yaxis = list(title = 'Value [%]')
    ) 
p

```



## Exercise: Plotting missing data 

Aim: Assess complete cases in the dataset and visualize missing data with Multivariate Imputation by Chained Equations (MICE) package. 

### Complete cases for heart rate

The function complete.cases() returns a logical vector indicating which cases are complete.Here we are using the negation (!) to indicate which cases are not complete.
    
   
```{r message=FALSE, warning=FALSE}

library(mice)

#install.packages("VIM")
library(VIM)

# list rows of data that have missing values 
hr[!complete.cases(hr$SUBJECT_ID),]

hr[!complete.cases(hr$VALUEUOM),]

```
We have all subjects IDs and heart rate units, as expected.


```{r message=FALSE, warning=FALSE}

library(mice)

#install.packages("VIM")
library(VIM)

# list rows of data that have missing values 
hr[!complete.cases(hr$VALUENUM),]

```

And we have only 1 patient with no heart rate value (correspondent to the outlier we previously removed).


### Plotting missing values with MICE package for heart rate

We observe that for the case of heart rate there are roughly 100% of values in the data set. 

```{r message=FALSE, warning=FALSE}

library(mice)

#install.packages("VIM")
library(VIM)

mice_plot_hr <- aggr(hr$VALUENUM, col=c('navyblue','#2980b9'), numbers=TRUE, sortVars=TRUE,
labels=names(hr$VALUENUM), cex.axis=.7, gap=3, ylab=c("Missing data","Pattern"))

```


### Omit cases for heart rate

The function na.omit() returns the object with listwise deletion of missing values.


```{r message=FALSE, warning=FALSE}

# create new dataset without missing data 
hr_complete <- na.omit(hr)


```
And now we have a complete dataset with missing data preprocessed for heart rate measurements. 


Let's repeat the steps for the variable pulse oximetry.

### Complete cases for pulse oximetry

```{r message=FALSE, warning=FALSE}

# list rows of data that have missing values 
sp[!complete.cases(sp$SUBJECT_ID),]

sp[!complete.cases(sp$VALUEUOM),]

```
We have all subjects IDs and spO2 units, as expected.


```{r message=FALSE, warning=FALSE}

# list rows of data that have missing values 
sp_incomplete <- sp[!complete.cases(sp$VALUENUM),]

```
And we have 2437 patients with no pulse oximetry value (part corresponds to the outliers we previously removed). All patients with outliers have now missing values, therefore their entries correspond to incomplete cases.


### Plotting missing values with MICE package for pulse oximetry

We observe that for the case of pulse oxymetry there are 91.4% values in the data set with no missing value. There are 8.6% missing values in the data. 

```{r message=FALSE, warning=FALSE}

library(mice)

#install.packages("VIM")
library(VIM)


mice_plot_sp <- aggr(sp$VALUENUM, col=c('navyblue','#2980b9'), numbers=TRUE, sortVars=TRUE,
labels=names(sp$VALUENUM), cex.axis=.7, gap=3, ylab=c("Missing data","Pattern"))

```


### Omit cases for pulse oximetry

The function na.omit() returns the object with listwise deletion of missing values.

```{r message=FALSE, warning=FALSE}

# create new dataset without missing data 
sp_complete <- na.omit(sp)


```
And now we have a complete dataset with missing data preprocessed for pulse oximetry measurements. Attention however, because missing data might not be missing at random (MNAR - see chapter contents for details) and in that case it might be important to identify and handle in a different way the missing values in the dataset. Eg. this can be done through missing data imputation.


We can now see how the distribution looks like for complete and processed data.

```{r message=FALSE, warning=FALSE}

#install.packages("plotly")

library(plotly)

p <- plot_ly(y = ~sp_complete$VALUENUM,  name = "Pulse oximery [%]", type = "box") %>%
  add_trace(y =~hr_complete$VALUENUM,  name = "Heart rate [bpm]", type = "box") %>%
layout(
    title = "Pulse oximetry and heart rate distribution for all patients",
    yaxis = list(title = 'Value')
    ) 
p

```

## Exercise: Missing data imputation

Aim: To impute missing data using several methods.

There are several approachs for missing data imputation. Here we show how to impute missing data with packages available in R studio.

Before missing data imputation we must first remove outliers, so our imputation is performed based on values within the pysiological ranges.

Let's select heart rate and pulse oximetry values in the same dataframe.

```{r message=FALSE, warning=FALSE}

hr_sp <- vitals[(((vitals$ITEMID = 211) | (vitals$ITEMID = 220045)) & (vitals$VALUEUOM == "bpm")) | (((vitals$ITEMID = 646) | (vitals$ITEMID = 220277)) & (vitals$VALUEUOM == "%")),]

```

And recode as "NA" where values are outside the physiological ranges.

```{r message=FALSE, warning=FALSE}

c = (((hr_sp$ITEMID = 211) | (hr_sp$ITEMID = 220045)) & ((hr_sp$VALUENUM > 300) | (hr_sp$VALUENUM < 0)))

hr_sp$VALUENUM[c] <- NA

c = (((hr_sp$ITEMID = 646) | (hr_sp$ITEMID = 220277)) & ((hr_sp$VALUENUM > 100) | (hr_sp$VALUENUM < 80)))

hr_sp$VALUENUM[c] <- NA


```


### Imputation of the population median value

We can directly impute the median of the population for each variable.

For heart rate:
```{r message=FALSE, warning=FALSE}

df = hr_sp

c = (((df$ITEMID = 211) | (df$ITEMID = 220045)) & (df$VALUEUOM == "bpm"))

#pop_hrmedian <- median(df[c,]$VALUENUM, na.rm=TRUE) 

df[c,]$VALUENUM[is.na(df[c,]$VALUENUM)] = median(df[c,]$VALUENUM, na.rm=TRUE) 

```

For pulse oximetry:
```{r message=FALSE, warning=FALSE}

df = hr_sp

c = (((df$ITEMID = 646) | (df$ITEMID = 220277)) & (df$VALUEUOM == "%"))

#pop_spmedian <- median(df[c,]$VALUENUM, na.rm=TRUE) 

df[c,]$VALUENUM[is.na(df[c,]$VALUENUM)] = median(df[c,]$VALUENUM, na.rm=TRUE)

```

### Imputation with the Amelia package

Let’s impute the missing values with Amelia package, named after Amelia Earhart.

We selected pulse oximetry and heart rate as example, however this can be performed for all the variables. We created two dataframes, one with information of heart rate measurements and the other with information regarding pulse oximetry measurements.

```{r message=FALSE, warning=FALSE}

df_hr = hr_sp[(((hr_sp$ITEMID = 211) | (hr_sp$ITEMID = 220045)) & (hr_sp$VALUEUOM == "bpm")),]

colnames(df_hr)[colnames(df_hr)=="VALUENUM"] <- "hr"


df_sp = hr_sp[(((hr_sp$ITEMID = 646) | (hr_sp$ITEMID = 220277)) & (hr_sp$VALUEUOM == "%")),]

colnames(df_sp)[colnames(df_sp)=="VALUENUM"] <- "sp"

```



Here we will select 1 patient ICU stay as example. Then we merge both dataframes and visualize missingness.

```{r message=FALSE, warning=FALSE}

icustay = 210989

df_hr = df_hr[df_hr$ICUSTAY_ID == icustay,c("ICUSTAY_ID","hr")]

df_sp = df_sp[df_sp$ICUSTAY_ID == icustay,c("ICUSTAY_ID","sp")]

df  <- merge(df_hr, df_sp, by = c("ICUSTAY_ID"))

library(Amelia)

missmap(df,
 main = "Missingness Map: heart rate (hr) & spO2 (sp)",
 y.labels = NULL,
 y.at = NULL)

```

Finally, we input missing data using Amelia package. For the cases (correspondent to rows) where there is no value for heart rate or pulse oxymetry, the value is not imputed. With a dataset containing higher amount of variables, we will not have so many of these cases. However, in the cases where we do not have any value we can proceed with eg. case deletion. 

```{r message=FALSE, warning=FALSE}

library(Amelia)

#specify columns and run amelia
amelia_fit <- amelia(df[,c(2,3)], parallel = "multicore")

```

### Analyze the imputed values

We will analyze the imputed values created to understand if the imputation method was adequate. A common practice consists in comparing the distribution of the imputed values and of the observed values. Check the compare.density function.

For heart rate:

```{r message=FALSE, warning=FALSE}

compare.density(amelia_fit, var = "hr")

```


For pulse oximetry:

```{r message=FALSE, warning=FALSE}

compare.density(amelia_fit, var = "sp")

```

We observe that other methods should be applied in order to have a more fitted distribution of the data.
