---
title: "Linear Regression Exercises"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description
This document contains the codes and scripts to follow the manuscript and contents for the Section 2.08.

## 3a: Introduction to Data Analysis Approaches

Import dataset from **PhysioNet**

```{r ImportData, echo=TRUE}
url <- "https://archive.physionet.org/physiobank/database/mimic2-iaccd/full_cohort_data.csv"
dat <- read.csv(url)
```
In case it is not rendering, try the following command.
```{r VariableNames, echo=TRUE}
dat <- read.csv("full_cohort_data.csv")
```
Can you notice what is the difference between the first chunk and the second one?
The first one works when you are online and get access to the content stored remotely. 

For the second one, you should have already dowloaded the csv file in the same folder that contains this Rmd file. For your convenience, this *.csv is available in the same remopository.

Once it is imported, let's take a look on the variables included in this dataset:

```{r VariableNames, echo=TRUE}
names(dat)
```
Retrieve the dimension (number of rows and columns) of *dat*:

```{r DataDimensions, echo=TRUE}
dim(dat)
```

## 3b: Linear Regression

Let's visualize the dataset with a scatter plot:

```{r pressure, echo=TRUE}
plot(dat$pco2_first,dat$tco2_first,xlab="PC02", ylab="TC02",pch=19,xlim=c(0,175))
```
Let's fit the data to a linear regression model:
```{r, echo=TRUE}
co2.lm <- lm(tco2_first~pco2_first,data=dat)
```
Let's display a summary of this fit:
```{r, echo=TRUE}
summary(co2.lm)
```
Fitting a quadratic model:
```{r, echo=TRUE}
co2.quad.lm <- lm(tco2_first~pco2_first + I(pco2_first^2),data = dat)
summary(co2.quad.lm)$coef
```
Let's best fit lines to the scatter plots using the **abline** function
```{r Fittings, echo=TRUE}
plot(dat$pco2_first,dat$tco2_first,xlab="PC02", ylab="TC02",pch=19,xlim=c(0,175))
abline(co2.lm,col='red')
abline(co2.quad.lm,col='blue')
```
## Statistical Interactions and Testing Nested Models
Check what type of variable is assuming RStudio for **gender_num**.
```{r, echo=TRUE}
class(dat$gender_num)
```
Set the correct class for **gender_num**:
```{r, echo=TRUE}
dat$gender_num <- as.factor(dat$gender_num)
class(dat$gender_num)
```
Fit again once it is encoded correctly **gender_num** and retrieve summary:
```{r, echo=TRUE}
co2.gender.lm <- lm(tco2_first~pco2_first+gender_num,data = dat)
summary(co2.gender.lm)$coef
```
Plot fitting lines and compare:
```{r, echo=TRUE}
plot(dat$pco2_first,dat$tco2_first,col=dat$gender_num,xlab = "PCO2", ylab = "TCO2", 
     xlim = c(0,40), type = "n", ylim = c(15,25))
abline(a = c(coef(co2.gender.lm)[1]), b = coef(co2.gender.lm)[2])
abline(a = coef(co2.gender.lm)[1]+coef(co2.gender.lm)[3],b=coef(co2.gender.lm)[2],col="red")
```
Modeling taking into consideration variables' interactions:
```{r, echo=TRUE}
co2.gender.inteaction.lm<-lm(tco2_first~pco2_first*gender_num,data = dat)
summary(co2.gender.inteaction.lm)$coef
```

```{r, echo=TRUE}
plot(dat$pco2_first,dat$tco2_first,col=dat$gender_num,xlab = "PCO2", ylab = "TCO2", 
     xlim = c(0,40), type = "n", ylim = c(15,25))
abline(a = c(coef(co2.gender.lm)[1]), b = coef(co2.gender.lm)[2])
abline(a = coef(co2.gender.lm)[1]+coef(co2.gender.lm)[3],b=coef(co2.gender.lm)[2],col="red")
abline(a=coef(co2.gender.inteaction.lm)[1],b=coef(co2.gender.inteaction.lm)[2],lty=3,lwd = 2)
abline(a=coef(co2.gender.inteaction.lm)[1] + coef(co2.gender.inteaction.lm)[3], 
       b = coef(co2.gender.inteaction.lm)[2]+coef(co2.gender.inteaction.lm)[4],
       col = "red", lty = 3, lwd = 2)
legend(24,20,lty = c(1,1,3,3),lwd = c(1,1,2,2), col = c("black","red","black","red"),
       c("Female","Male","Female (Interaction Model","Male (Interaction Model)"))
```

Perform anova analysis:
```{r, echo=TRUE}
anova(co2.lm,co2.gender.inteaction.lm)
```

## Confidence and Prediction Intervals
Get confidence intervals:
```{r, echo=TRUE}
confint(co2.lm)
```

Predict the outcome over the range of covariate values we observed determined by the min and max functions:
```{r, echo=TRUE}
grid.pred<-data.frame(pco2_first=seq.int(from=min(dat$pco2_first,na.rm = T),
                                         to=max(dat$pco2_first,na.rm = T)))
preds<-predict(co2.lm,newdata = grid.pred,interval = "prediction")
preds[1:2,]
```

```{r, echo=TRUE}
plot(dat$pco2_first,dat$tco2_first,xlab = "PCO2",ylab = "TCO2",pch=19,xlim = c(0,175))
co2.lm <- lm(tco2_first ~ pco2_first, data = dat)
abline(co2.lm,col="red",lwd=2)
lines(grid.pred$pco2_first,preds[,2],lty=3)
lines(grid.pred$pco2_first,preds[,3],lty=3)
```
